% \chapter*{Preface}

% Significant parts of this dissertation are based on papers published previously.  In particular: \Cref{chap:tsp}, \nameref{chap:tsp}, is based on \cite{Cooper2019}.  \Cref{chap:gcol}, \nameref{chap:gcol}, is based on \cite{Cooper2019a}.  \Cref{chap:median}, \nameref{chap:median}, presents work from \cite{Cooper2018} and \cite{Cooper2021}.  \Cref{chap:nmp}, \nameref{chap:nmp}, presents work which appeared in an early form in \cite{Cooper2020}, and which has been significantly revised and is under review with the Journal of Membrane Computing as at the time of submission of this dissertation.

% \section*{Acknowledgements}
\chapter*{Acknowledgements}

Acknowledgement must be made, firstly, of my supervisors, Doctor Radu Nicolescu and Associate Professor Patrice Delmas.  Without their support and guidance, I certainly would have never made it to this point.  So too must acknowledgement be made of Professor Emeritus Georgy Gimel'farb and Doctor Wannes van der Mark, both of whom have provided invaluable support at times.    Also José Calderón Trilla.

Furthermore, thanks must be given to the professional staff from: the School of Computer Science; the School of Graduate Studies; Libraries and Learning Services (particularly those who served as Computer Science subject librarian); the Centre for eResearch; and the wider University of Auckland.  Particular mention must be made of Kaleigh Fennah, Emma Gavenda, Sue Skelly, Sarah Sneyd and Robyn Young.

I also would like to mention several people who were students contemporaneously with me (and in a few cases were also co-authors with me), including: Arabella Anderson; Mihailo Azhar; Daniel Britten; Doctor Trevor Gee; Alec Henderson; Yan Kolezhitskiy; Yezhou Liu; William Hsu; and Eirian Perkins.  Each of these people in some way has improved my time and experience during the PhD process.

I'm fairly sure there are further people who should be mentioned, but I'm afraid that I can't recall them specifically at the time of writing.

% Both the University of Auckland overall, and especially the School of Computer Science specifically, should also be thanked for their financial support.  The University provided me with a scholarship \(3\frac{1}{2}\) years, and the School provided a further six-month top-up.  I would have been unable to undertake this degree without said financial support.

% Everyone who assisted me with the use of the languages in [choiceoflang chapter], especially those people who responded to my requests for assistance on the MLton and Guile mailing lists, and in the Racket Slack workspace.  Special mention must be made of: Yawar Raza from the Rochester Institute of Technology, who routinely answered my questions about MLton very quickly and helpfully; and [soegaard2 from Racket slack]

% Lastly, and above all else, mention must be made of my mum.  Thanks mum!

\vspace{1cm}
\hfill James Cooper

\hfill Auckland, New Zealand

\hfill September 2021

