\newcommand{\bo}{\(b\)}

\chapter{\label{chap:gcol}The Graph Colouring Problem}

\section{\label{sec:gcol:background}Background}

The \gls{gcp} is a deceptively simple problem in graph theory.  In its most basic form, it is the problem of assigning labels such as colours to the nodes in a graph so that no two nodes connected by an edge share a colour, usually with the addition of the extra requirement either of finding the minimum number of colours required or only using a number of colours up to some upper bound --- of course, for any graph with \(N\) nodes, it is trivial to colour it as required using \(N\) colours.  The problem finds applications in many areas, including timetabling, register allocation for compilers, as well as solving Sudoku puzzles (which are in turn a special case of Latin Squares) \cite{Lewis2016}.  The general case is NP-Complete, and thus no polynomial time solution has been found so far (finding one would prove that P = NP), though they have been found for specific forms of graph or particular numbers of colours.  For example, in the case of 3-colouring, the current best known solution takes \bigoh{1.3289^n} time \cite{Beigel2005}.

% Gheorghe \textit{et al.} presented a \gls{ps} solution to the 3-colouring problem, using communicating \gls{skps} \cite{Gheorghe2013}.  Based on this work, we first discuss in \cref{sec:gcol:cml} a \gls{cml} implementation of the \gls{skps} solution in \cite{Gheorghe2013}, briefly compare the two approaches and indicate some other \gls{ps} variants we think \gls{cml} might be a good fit with.  We then present in \cref{sec:gcol:cpsys} a fairly concise single-cell solution to the problem using \gls{ps} with compound terms (\gls{cps} for short -- see \cite{Nicolescu2018} for a detailed background on \gls{cps}), and provide an informal analysis of it.  Lastly, in \cref{sec:gcol:examples}, we provide examples of the operation of our cP~system on the 3-colour problem, both for a graph that is colourable and one that is not, before concluding in \cref{sec:gcol:conc}.

% To the best of our knowledge, this is the first time \gls{cps} has been applied to graph colouring, and the first time that a \gls{cml} implementation has been used to simulate the operation of any P~system.

% We wish to emphasise that we do not consider our simulation or \gls{cps} solution to be `superior' to that of Gheorghe \textit{et al.} (though we do think it demonstrates the utility of \gls{cps}), but instead wish to complement theirs and others' work with another alternative.

% \input{chapters/gcol/tex/cml}
% \input{chapters/gcol/tex/cprules}
% \input{chapters/gcol/tex/examples}

% \section{\label{sec:gcol:conc}Conclusions}
% Firstly we implemented a version of Gheorghe \textit{et al.}'s \gls{skps} solution to the 3-colouring problem using \fsharp{} and a \gls{cml}-derived library, Hopac.   On the small sample of graphs tested this produces reasonable results for smaller graphs, at least as good as those achieved by the P-Lingua/MeCoSim simulation.  In both systems, the inclusion of a test for already-invalid compartments has the potential to improve the runtimes significantly, as it prunes the search space, sometimes quite dramatically.

% \gls{cml} appears to be a good fit in principle to \gls{ps} variants that use communication between cells/membranes, but this particular problem had fairly minimal communication and instead relied heavily upon cell division.  Further testing with other problems is required to determine the efficacy of \gls{cml} for sure.

% We secondly provided a single-top-level-cell \gls{cps} solution to the \gls{gcp}, which is capable of solving the problem for arbitrary graphs with an arbitrary number of potential colours in at most \(|V| + 1\) steps.  Its operation was demonstrated with two 3-colouring examples based on two different graphs, for which it was and was not possible, respectively, to find a valid 3-colouring.

\citeauthor{Gheorghe2013} presented a \gls{ps} solution to the 3-colouring problem, using communicating \gls{skps} \cite{Gheorghe2013}.  Based on this work, this \namecref{chap:gcol} first discusses in \cref{sec:gcol:cml} a \gls{cml} implementation of the \gls{skps} solution in \cite{Gheorghe2013}, briefly compare the two approaches and indicate some other \gls{ps} variants which \gls{cml} might be a good fit with.  This \namecref{chap:gcol} then presents in \cref{sec:gcol:cpsys} a concise single-cell solution to the problem using \gls{cps}, and provide an informal analysis of it.  Lastly, in \cref{sec:gcol:examples}, this \namecref{chap:gcol} provides examples of the operation of our cP~system on the 3-colour problem, both for a graph that is colourable and one that is not, before concluding in \cref{sec:gcol:conc}.

% So far as known, this is the first time \gls{cps} has been applied to graph colouring, and the first time that a \gls{cml} implementation has been used to simulate the operation of any P~system.

\input{chapters/gcol/tex/cml}
\input{chapters/gcol/tex/cprules}
\input{chapters/gcol/tex/examples}

\section{\label{sec:gcol:conc}Conclusions}
Firstly, this \namecref{chap:gcol} implemented a version of \citeauthor{Gheorghe2013}'s \gls{skps} solution to the 3-colouring problem using \fsharp{} and a \gls{cml}-derived library, Hopac.   On the small sample of graphs tested, this produces reasonable results for smaller graphs, at least as good as those achieved by the \gls{plingua}/\gls{mecosim} simulation.  In both systems, the inclusion of a test for already-invalid compartments has the potential to improve the runtimes significantly, as it prunes the search space, sometimes quite dramatically.

\gls{cml} appears to be a good fit in principle to \gls{ps} variants that use communication betthis \namecref{chap:gcol}en cells/membranes, but this particular problem had fairly minimal communication and instead relied heavily upon cell division.  Further testing with other problems is required to determine the efficacy of \gls{cml} for sure.

This \namecref{chap:gcol} secondly provided a single-top-level-cell \gls{cps} solution to the \gls{gcp}, which is capable of solving the problem for arbitrary graphs with an arbitrary number of potential colours in at most \(|V| + 1\) steps.  Its operation was demonstrated with two 3-colouring examples based on two different graphs, for which it was and was not possible, respectively, to find a valid 3-colouring.
